RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-09-06 doc/bioinformatics2013/RVD2.tex
Finished bayesian hyp test section.

Since, each position $j$ is exchangeable given the genome-wide parameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors which increases the speed considerably.

%%%%%%%%%%%%%%%%%%
% Posterior Density Test
%%%%%%%%%%%%%%%%%%
\subsection{Bayesian hypothesis test for high error rate}

To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
\begin{align}
 H_1: \mu_{case}-\mu_{control}\geq\tau \\
 H_2: \mu_{case}-\mu_{control}<\tau
\end{align}

We can calculate the posterior probability of the hypothesis $H_1$ being true by integrating the posterior density over the correct region:

\begin{equation}
 p(H1:\mu_{case}-\mu_{control}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
\end{equation}

We accept the hypothesis $H_1$ if this posterior probability is no lower than 95\%, which means we are 95\% sure that there is a mutation in this position.

ROC curve is plotted to evaluate the performance of the detection algorithm with different threshold for different mutation rate.

\subsection{$\chi^2$ test for non-uniform base distribution}

>>>>

Since, each position $j$ is exchangeable given the genome-wide parameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors which increases the speed considerably. We have found that the algorithm performance improves when we take the median of ten or more M-H samples as a single Gibbs step for each position. This computational effort only minimally increases run time when run on a parallel computing platform. In the experiments in the remainder, we use a XX sample burn-in to allow the Markov chain to approach its stationary distribution and we thin the chain by a factor of 2 to reduce autocorrelation among samples.

%%%%%%%%%%%%%%%%%%
% Posterior Density Test
%%%%%%%%%%%%%%%%%%
\subsection{Bayesian hypothesis test for high error rate}
The Metropolis-within-Gibbs (M-w-G) sampling algorithm provides samples from the posterior distribution of $\mu_j$ for both the control and case data. For notational simplicity in this section, we define the random variables associated with these two distributions $\mu_j^{\text{case}}$ and $\mu_j^{\text{control}}$.

A variant is called if the data indicate that $\mu_j^{\text{case}} > \mu_j^{\text{control}}$ with high confidence,
\begin{equation}
	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} \geq \tau ) > 1-\alpha,
\end{equation}
where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level. While analytical forms for the posterior distributions are not available, hthe M-w-G algorithm does provide samples from those distributions. So, we sample from the posterior distribution $\mu_j^{\Delta} \triangleq \mu_j^{\text{case}} - \mu_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\mu_j^{\text{case}}$ and$\mu_j^{\text{control}}$ and calculating samples from $\mu_j^{\Delta}$ directly.

%%%%%%%%%%%%%%%%%%
% Chi^2 Test
%%%%%%%%%%%%%%%%%%
\subsection{$\chi^2$ test for non-uniform base distribution}
